Fail Faster

### 2025-03-16 paper/paper.tex
Changes on Zoom with Thia

\newif\ifdraft\drafttrue{}
\newif\iflater\latertrue{}

@@

%% Rights management information.  This information is sent to you
%% when you complete the rights form.  These commands have SAMPLE
%% values in them; it is your responsibility as an author to replace
%% the commands and values with those provided to you when you
%% complete the rights form.
\setcopyright{acmlicensed}
\copyrightyear{2025}
\acmYear{2025}
\acmDOI{XXXXXXX.XXXXXXX}
%% These commands are for a PROCEEDINGS abstract or paper.
\acmConference[Conference acronym 'XX]{Make sure to enter the correct
  conference title from your rights confirmation email}{June 03--05,
  2018}{Woodstock, NY}
%%
%%  Uncomment \acmBooktitle if the title of the proceedings is different
%%  from ``Proceedings of ...''!
%%
%%\acmBooktitle{Woodstock '18: ACM Symposium on Neural Gaze Detection,
%%  June 03--05, 2018, Woodstock, NY}
\acmISBN{978-1-4503-XXXX-X/2025/02}

@@

%%
%% The "title" command has an optional parameter,
%% allowing the author to define a "short title" to be used in page headers.
\title{Fail Faster}

@@

%%
%% The abstract is a short summary of the work to be presented in the
%% article.
\begin{abstract} \tr{Abstract needs a lot of work.}
Property-based testing (PBT) is a software testing approach that verifies a system against a large suite of automatically-generated inputs. Since testing pipelines typically run under strict time constraints, PBT generators must produce inputs as quickly as possible to maximize the likelihood of finding bugs in the time available. However, existing PBT libraries often prioritize generality at the cost of performance. We introduce \texttt{waffle\_house}, a high-performance generator library that uses staging, a lightweight compilation technique, to eliminate many common generator abstractions. To evaluate \texttt{waffle\_house}, we design a novel benchmarking methodology that compares generators based on program equivalence, isolating performance improvements from differences in input distribution. Using this methodology, we compare \texttt{waffle\_house} to a leading generator library, and, through extensive evaluation over a diverse range of generators, demonstrate that \texttt{waffle\_house} significantly improves generation speed and resource efficiency while matching \texttt{base\_quickcheck}'s expressiveness.
\end{abstract}

@@

%%
%% Keywords. The author(s) should pick words that accurately describe
%% the work being presented. Separate the keywords with commas.
\keywords{Property-based testing, Generators, Staging, Meta-programming}
%% A "teaser" image appears between the author and affiliation
%% information and the body of the document, and typically spans the
%% page.

@@

Property-based testing (PBT) is a technique in which a \textit{generator}
produces random inputs to a system under test, which is then evaluated against a
set of properties the system is expected to satisfy.  Recent studies on PBT
usage have shown that practitioners often run tests with very short time
budgets --- typically between 50 milliseconds and 30 seconds \cite{inpractice} \jwc{phrasing could suggest ``race against time'' aspect.}. Given this
constraint, generating inputs as quickly as possible is crucial for maximizing
the odds of finding a bug in the limited time available. In other words,
generators must be treated as performance-sensitive code.
A high-leverage way to improve the performance of generators is to optimize 
\textit{generator libraries}: if
programmers are provided with the tools to write fast generators, then they 
do not themselves need to be performance engineers. \jwc{... to find bugs fast}

However, the performance of 
existing generator libraries falls short of what is possible. This is due to two major 
sources of inefficiency. First, the same high-level design that gives generator libraries their flexibility and expressiveness also introduces overhead:
layered abstractions and indirections hinder the compiler's ability to optimize computations, shifting more work to runtime. \jwc{Nice!!}
Second, the process of 
selecting random \proposechange{values}{numbers} \jwc{We should reserve ``values'' for BSTs/etc... otherwise this sentence is confusing. What else does would a generator's runtime consist of, except for generating values?}(``random sampling'') can constitute an unexpectedly high proportion
of a generator's runtime, depending on one's choice of \rand and
the number of samples needed. This \jwc{Unclear referent here} raises a fundamental 
question: can generator libraries produce code that is both expressive 
\textit{and} efficient?
\jwc{Love these first two paragraphs.}
%In practice, PBT is used 
%as a lightweight, ad hoc approach to software correctness,
%\jwc{Move this bit about how it's used to the first sentence. I.e. PBT is a testing technqique where random data is fed to a system under test...}
%aimed at 
%detecting bugs within a short testing window---typically 50 milliseconds 
%to 30 seconds. \jwc{This sells us a little short: we should say something to the effect of ``Recent work on understanding how PBT is used in practice has determined that practitioners often run PBT with very short time budgets...''} Given this practical constraint, it is imperative that 
%inputs be generated as fast as possible; in other words, generators must be 
%treated as performance-sensitive code.

@@

  % in strict functionallanguages like OCaml and Scal
To answer this, we propose a two-pronged approach for designing generator libraries 
that preserves their expressiveness while eliminating unnecessary overhead.
First, we describe a method of optimizing
generator libraries using \textit{multi-stage programming}, or ``staging'', which has 
long been used for optimizing idiomatic functional DSLs \cn.
We extend this technique to generators, completely eliminating the overhead of many common generator abstractions. For example, in unstaged generator libraries, each call to monadic bind constructs and immediately executes a closure at runtime \jwc{We should probably say ``usually'' here... in very simple scenarios the compiler can figure it out}.
\jwc{Todo: joe --- cite some evidence for this}
Meanwhile, a staged approach can resolve the bind at compile time, enabling the compiler to inline and eliminate these closures.
Secondly, we demonstrate the
impact of different choices of \rand on generator performance 
by conducting a controlled comparison. Together, these insights provide a
recipe for improving even state-of-the-art generator libraries.
\jwc{Great, like this a lot.}

@@

To demonstrate the effectiveness of our technique, we apply it to \bq, the fastest property-based testing library in OCaml according to our tests.
%\jwc{Unclear if this is true, I think we should just say that by our testing it's the fastest in OCaml.}
Our staged version of this library,\\ \name, is a direct drop-in replacement for \bq that 
is \textit{completely semantically equivalent}: that is, given identical random seeds, generators written using \name{}
produce exactly the same values as those written in \bq. However, in \name, 
the abstractions provided by the generator DSL are zero-cost, and the \rand is 
highly-optimized. The direct equivalence between \name and \bq allows us to 
perform extremely fine-grained \propose{bug-finding} comparisons, as generators 
implemented using both versions of the library will produce exactly the same 
sequence of values. Therefore, differences in \proposechange{performance}{bug-finding ability} are directly 
attributable to \name's \proposechange{speedup}{??}.

To evaluate \name's effectiveness \jwc{This could flow better from the end of the last para --- we were just talking about evaluating \name{}'s effectivness.}, we
implement a series of workloads \jwc{this is an etna term of art} in each generator library, and show that 
\name runs faster \proposecut{and allocates less} than \\\bq. Further, we show that this performance benefit
translates to bug-finding ability by running these workloads in Etna, a platform that 
simulates real-world PBT usage by injecting bugs into a system and measuring how quickly various 
generators detect them.
Finally, to demonstrate the generality of our technique, we implement \name in Scala
and benchmark the same generators in the Scala port of the library---thereby highlighting 
its portability
to other strict functional languages.
\jwc{I think we should more explicitly point out the portability earlier, maybe up where we say we're applying it to BQ.}

>>>>

\newif\ifdraft\drafttrue{}
\newif\iflater\laterfalse{}

@@

%% Rights management information.  This information is sent to you
%% when you complete the rights form.  These commands have SAMPLE
%% values in them; it is your responsibility as an author to replace
%% the commands and values with those provided to you when you
%% complete the rights form.
\setcopyright{none}
% \copyrightyear{2025}
% \acmYear{2025}
% \acmDOI{XXXXXXX.XXXXXXX}
%% These commands are for a PROCEEDINGS abstract or paper.
% \acmConference[Conference acronym 'XX]{Make sure to enter the correct
%   conference title from your rights confirmation email}{June 03--05,
%   2018}{Woodstock, NY}
%%
%%  Uncomment \acmBooktitle if the title of the proceedings is different
%%  from ``Proceedings of ...''!
%%
%%\acmBooktitle{Woodstock '18: ACM Symposium on Neural Gaze Detection,
%%  June 03--05, 2018, Woodstock, NY}
% \acmISBN{978-1-4503-XXXX-X/2025/02}

@@

%%
%% The "title" command has an optional parameter,
%% allowing the author to define a "short title" to be used in page headers.
\title{Fail Faster}
\subtitle{\hg{TODO: Subtitle}}

@@

%%
%% The abstract is a short summary of the work to be presented in the
%% article.
\begin{abstract}
  \hg{TODO: Abstract}
\end{abstract}

@@

%%
%% Keywords. The author(s) should pick words that accurately describe
%% the work being presented. Separate the keywords with commas.
% \keywords{Property-based testing, Generators, Staging, Meta-programming}
%% A "teaser" image appears between the author and affiliation
%% information and the body of the document, and typically spans the
%% page.

@@

Property-based testing (PBT) is a technique for software testing which uses
random test inputs to evaluate logical properties that the system is expected to
satisfy.  Recent studies on PBT usage~\cite{inpractice} have shown that
practitioners interleave PBT with their software development process, sometimes
testing their properties every time they save their code. This workflow makes
sense because the developer who just wrote the code is often the best person to
fix a potental bug; if a bug is found later, it will need to be fixed by someone
with less context.  All of this means that PBT is often run with very short time
budgets --- e.g., between 50 milliseconds and 30 seconds --- and it is important
that bugs can be found in that time-frame.
\jwc{phrasing could suggest ``race against time'' aspect.}\hg{Agree, can't see
how to fit this in yet}

Given these strict timelines, it is important to think of PBT as
performance-sensitive process. One important locus for performance improvement
is the \textit{generators} that produce random inputs to the properties under
test. These generators are written with the help of \textit{generator
libraries}, usually expressed as embedded domain-specific languages (DSLs) that
provide combinators for building and composing generators. Optimizing these
libraries is a key opportunity to speed up PBT: if programmers are
provided with the tools to write fast generators, then they do not themselves
need to be experts in performance tuning. \jwc{... to find bugs fast}

We find that the performance of existing generator libraries falls short of what
is possible. This is due to two major sources of inefficiency.
First, the same high-level design that gives generator libraries their
flexibility and expressiveness also introduces overhead: layered abstractions
and indirections hinder the compiler's optimizations, shifting more work to
runtime.
Second, the \rand that underlies the generator's operation --- the source of
random numbers that are used to make choices in the generator --- 
can constitute an unexpectedly high proportion of a generator's runtime.
\iflater
\hg{TODO later, I'm not 100\% sure it's clear how this particular question is
raised by the above. We could strengthen that connection}
\fi
These observations together raise a fundamental question: can generator
libraries produce code that is both expressive \textit{and} efficient?
%In practice, PBT is used 
%as a lightweight, ad hoc approach to software correctness,
%\jwc{Move this bit about how it's used to the first sentence. I.e. PBT is a testing technqique where random data is fed to a system under test...}
%aimed at 
%detecting bugs within a short testing window---typically 50 milliseconds 
%to 30 seconds. \jwc{This sells us a little short: we should say something to the effect of ``Recent work on understanding how PBT is used in practice has determined that practitioners often run PBT with very short time budgets...''} Given this practical constraint, it is imperative that 
%inputs be generated as fast as possible; in other words, generators must be 
%treated as performance-sensitive code.

@@

  % in strict functionallanguages like OCaml and Scal
To address this question, we propose a two-pronged approach for designing
generator libraries that preserves their expressiveness while eliminating
unnecessary overhead.
First, we describe a method of optimizing generator libraries using
\textit{multi-stage programming}, or ``staging,'' which has long been used for
optimizing idiomatic functional DSLs~\cn.
We extend this technique to generators, completely eliminating the overhead of
many common generator abstractions. For example, in unstaged generator
libraries, each call to monadic bind usually spends time constructing and
immediately executing a closure at runtime. \jwc{Todo: joe --- cite some evidence for this}
Meanwhile, a staged approach can resolve the bind at compile time, enabling the
compiler to inline and eliminate these closures.
Secondly, we demonstrate the impact of different choices of \rand on generator
performance by conducting a controlled comparison. \hg{More words here for
parallelism?} Together, these insights provide a recipe for improving even
state-of-the-art generator libraries.

@@

To demonstrate the effectiveness of our technique, we apply it to \bq, a
high-quality PBT library in OCaml, authored by Jane Street.
Our improved version, \name, is a direct drop-in replacement for
\bq that 
is \textit{completely semantically equivalent}: that is, given identical random seeds, generators written using \name{}
produce exactly the same sequence of values as those written in \bq. However, in \name, 
the abstractions provided by the generator DSL are zero-cost, and the \rand is 
highly-optimized. The direct equivalence between \name and \bq allows us to 
perform extremely fine-grained comparisons, differences in generation speed are
directly attributable to the staging and \rand optimizations.

We evaluate the optimizations in \name with a series of case studies in each
generator library, and show that \name runs faster than \bq.
Further, we show that this performance benefit\hg{wc?} translates to bug-finding
ability by running these case studies in the Etna framework~\cite{etna}, a
platform that simulates real-world PBT usage\hg{Don't love this phrase, I'll
think more about how we should talk about Etna} by injecting bugs into a system and
measuring how quickly various generators detect them.
Finally, to demonstrate the generality of our technique, we implement a version
of \name in Scala \hg{(based on scalacheck?)} and benchmark the same generators
in the Scala port of the library---thereby highlighting its portability to other
strict functional languages.
\iflater
\jwc{I think we should more explicitly point out the portability earlier, maybe
up where we say we're applying it to BQ.}\hg{Agree}
\fi
